\mytitle

### 2015-08-06 MSc_diss.tex
Finished objective and aims

Unlike many other air pollutants, tropospheric ozone is not directly emitted, but produced following the oxidation of carbon monoxide ($CO$), methane ($CH_4$), and nonmethane volatile compounds (nmVOCs) in the presence of nitrogen oxides (NOx) \citep{Crutzen1973,Myhre2013}. The emissions of these precursors arise from both natural and anthropogenic sources which have intensified since pre-industrial times \citep{Parrish2014,Volz1988} (Figure \ref{fig:O3observations}). Presumably it has led to roughly a doubling in background tropospheric ozone concentration \citep{Guicherit2000,Hartmann2013,Tarasick2008,Vingarzan2004}, although contribution from another major source, transport from the stratosphere, may also took place \citep{Fowler2008}. Climate is also believed to be a key driver of ozone production and destruction since the chemical reactions affecting ozone concentration are strongly dependent on such climatic factors as temperature, rainfall and humidity. Therefore in a changing climate tropospheric ozone is not anymore a local air quality issue, but is a global pollution problem \citep{Fowler2008}.

@@

According to the Intergovernmental Panel on Climate Change (IPCC) Fifth Assessment report (AR5) for present conditions (around year 2000) the global mean tropospheric ozone budget is approximately 331 Tg \citep{Myhre2013}. However, despite general agreement on how the drivers impact ozone concentrations at global and regional scales, its budget varies considerably between different modelling and observational studies \citep{Stevenson2006,Wild2007,Young2012}. Figure \ref{fig:O3budget} demonstrates that the range of the global tropospheric ozone budget estimates is approximately 60 Tg which is 18\% of the mean. In terms of ozone production the variation is even higher and accounts for 28\% of the corresponding value. In that regard, two questions arise: why does the tropospheric ozone budget differ between models and observations? And secondly, why does it vary to this extent, especially in terms of production? The only way to answer these questions is to perform a rigorous investigation of the factors that drive tropospheric ozone and attribute those differences to respective factors \citep{Young2012}.

\begin{figure}[t] % O3 budget
\centering
\includegraphics[width=0.7\linewidth]{o3_studies.eps}
\caption{Summary of tropospheric ozone global budget model and observation estimates for present (about 2000) conditions. All uncertainties quoted as 1 standard deviation (68\% confidence interval). Adapted from \citep{Myhre2013}}\label{fig:O3budget}
\end{figure}

A considerable effort has been made to tackle these problems, and generally there are two possible ways of doing it. The first one implies the direct analysis of tropospheric chemical schemes used in climate-chemistry models. For example, Emmerson and Evans (2009) compared the state of the art Master Chemical Mechanism which contains approximately 5600 species and 13500 reactions \citep{Jenkin2002} with other six chemical schemes and found four significant variations between them. This approach is linked with big challenges since a modern global model is a very sophisticated representation of the climate-chemistry interactions, and therefore it is often difficult to separate influence of one reaction from another. However, even being so developed climate-chemistry models do not take into account all known chemical reactions taking place in the real troposphere. A complete representation requires many thousands of species and tens of thousands of reactions, which is beyond the numerical capabilities currently available. This significantly limits the practicable size of chemical schemes and typically involves a reduction of the number of VOCs considered and lumping the carbon from the discarded species into representative surrogates.

! add something from report!

The second way to understanding the discrepancies between models and observations is to conduct an idealised theoretical study using a conceptual model. The advantage of this approach is that such kind of models can be quite easily constructed, and more importantly can be focused on representation of a certain chemical reactions which presumably could explain the discrepancy between models and observations.

\subsubsection*{Ozone tropospheric chemistry}\label{intro_O3chem}
%\input{O3CycleScheme}
Having identified the role of tropospheric ozone, it is time to remind ourselves the chemistry that controls abundance of it.

The key chemical processes that lead to $O_3$ formation and destruction in the troposphere are driven by reaction cycles involving free-radical intermediates, which are mainly formed from the photolysis of $O_3$ itself (\ref{reac:O3+hv=O1D+O2}) \citep{Fowler2008}. At UV wavelengths shorter than 320 nm, $O_3$ photolysis generates electronically excited oxygen atoms ($O(^1D)$), which than can either react with water vapour to form $OH$ radicals (\ref{reac:O1D+H2O=OH+OH}), or collide with an inert molecule (denoted '$M$') to reform $O_3$ (\ref{reac:O1D=O3P} and \ref{reac:O3P+O2=O3}):
\begin{equation}\label{reac:O3+hv=O1D+O2}
O_3 + hv \rightarrow O(^1D) + O_2
\end{equation}
\begin{equation}\label{reac:O1D+H2O=OH+OH}
O(^1D) + H_2O \rightarrow OH + OH
\end{equation}
\begin{equation}\label{reac:O1D=O3P}
O(^1D) \xrightarrow{M} O(^3P)
\end{equation}
\begin{equation}\label{reac:O3P+O2=O3}
O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
Once formed the OH radicals mainly react with $CO$ and $CH_4$ and initiate the $O_3$ formation or removal cycles by producing the peroxy radicals ($HO_2$, $CH_3O_2$):
\begin{equation}\label{reac:OH+CO=H+CO2}
OH + CO \rightarrow H + CO_2
\end{equation}
\begin{equation}\label{reac:H+O2=HO2}
H + O_2 \xrightarrow{M} HO_2
\end{equation}
\begin{equation}\label{reac:OH+CH4=CH3+H2O}
OH + CH_4 \rightarrow CH_3 + H_2O
\end{equation}
\begin{equation}\label{reac:CH3O+O2=CH3O2}
CH_3 + O_2 \xrightarrow{M} CH_3O_2
\end{equation}
The fate of the peroxy radicals depends on the presence of $NO_x$ ($=NO + NO_2$). In low-$NO_x$ conditions (remote regions), the $HO_2$ radical reacts with another $HO_2$ to form hydrogen peroxide ($H_2O_2$) or reacts with the methyl peroxy radical ($CH_3O_2$) to form methyl hydrogen peroxide ($CH_3OOH$):
\begin{equation} \label{reac:HO2+HO2=H2O2+O2}
HO_2 + HO_2 \rightarrow H_2O_2 + O_2
\end{equation}
\begin{equation} \label{reac:HO2+CH3O2=CH3OOH+O2}
HO_2 + CH_3O_2 \rightarrow CH_3OOH + O_2
\end{equation}
On a whole, in low-$NO_x$ conditions, a net $O_3$ destruction takes place, because the reaction sequence is initiated by $O_3$ photolysis. Plus in the same conditions some additional $O_3$ removal can occur because of the reaction of the $HO_2$ radical with $O_3$ leading to further destruction of ozone in a chain sequence involving formation of $OH$ radicals:
\begin{equation}\label{reac:HO2+O3=OH+2O2}
HO_2 + O_3 \rightarrow OH + 2O_2
\end{equation}
\begin{equation}\label{reac:OH+O3=HO2+O2}
OH + O_3 \rightarrow HO_2 + O_2
\end{equation}

In intermediate $NO_x$ conditions (rural areas of most industrialised countries), the peroxy radicals react with nitrogen monoxide ($NO$) to form nitrogen dioxide ($NO_2$), which subsequent photolysis leads to $O_3$ generation:
\begin{equation}\label{reac:HO2+NO=NO2+OH}
HO_2 + NO \rightarrow NO_2 + OH
\end{equation}
\begin{equation}\label{reac:CH3O2+NO=NO2+CH3O}
CH_3O_2 + NO \rightarrow NO_2 + CH_3O
\end{equation}
\begin{equation}\label{reac:NO2+hv=NO+O3P}
NO_2 + hv \rightarrow NO + O(^3P)
\end{equation}
\begin{equation}
O(^3P) + O_2 \xrightarrow{M} O_3\tag{\ref{reac:O3P+O2=O3}}
\end{equation}
As shown in Figure \ref{fig:O3scheme} reactions (\ref{reac:HO2+NO=NO2+OH}) and (\ref{reac:CH3O2+NO=NO2+CH3O}) form parts of the free-radical propagated $O_3$ forming cycles, which may occur a number of times before being halted by radical termination reactions (\ref{reac:HO2+HO2=H2O2+O2} and \ref{reac:HO2+CH3O2=CH3OOH+O2}), and the higher $NO_x$ level is the greater number of these cycles can occur. Therefore, in intermediate $NO_x$ conditions a net $O_3$ formation takes place and it is sensitive to $NO_x$.

@@

In high-$NO_x$ conditions (urban environment), the $OH$ radical may have another fate rather than reacting with $CO$, $CH_4$ and $O_3$. Depending on the concentration of $NO_2$ it can react with it to form nitric acid ($HNO_3$):
\begin{equation}\label{reac:OH+NO2=HNO3}
OH + NO_2 \xrightarrow{M} HNO_3
\end{equation}
The formation of $HNO_3$ terminates free-radical propagated $O_3$ forming cycles leading to a decrease in $O_3$ production rate. However, elevated emissions of $CO$ and $CH_4$ (and other nonmethane volatile organic compounds (nmVOCs)) may make reactions (\ref{reac:OH+CO=H+CO2})-(\ref{reac:OH+CH4=CH3+H2O}) more competitive with the reaction (\ref{reac:OH+NO2=HNO3}) and lead to an increase in the $O_3$ production rate. Therefore, in high-$NO_x$ conditions the $O_3$ production rate can either decrease or increase depending on $NO_x$ and $VOCs$ levels.

@@

therefore also having an impact on the rates of $O_3$ production \citep{Fowler2008} as do some other species, which are not mentioned in this section. Their role is described in other sections as far as it is necessary.

\subsection{Alkyl nitrates}\label{sec:intro_AN}
Alkyl nitrates ($RONO_2$) are organic trace gases, which through one of their formation mechanisms are closely linked with the production of tropospheric ozone \citep{Reeves2007}. This mechanism is present in the following section.
\subsubsection*{Alkyl nitrates chemistry}\label{into_ANchem}
The photochemical formation of the alkyl nitrates starts when chemistry of the troposphere reaches the regime of $O_3$ production, i.e. when reactions of the peroxy radicals with $NO$ start to compete with their self- and cross-reactions more effectively due to an increase in $NO_x$ mixing ratio. In this case the reaction between the alkyl peroxy radical ($RO_2$, where $R$ denotes methyl group, $CH_3$) and $NO$ can reveal its two-channel nature \citep{Day2003}:
\begin{subequations} \label{reac:RONO2alpha}
\begin{align}
RO_2 + NO &\xrightarrow{1-\alpha_3} RO + NO_2 \label{reac:RO2+NO=aphaRO+NO2}\\
RO_2 + NO &\xrightarrow{\alpha_3} RONO_2 \label{reac:RO2+NO=aphaRONO2}
\end{align}
\end{subequations}
The latter reaction can serve as a termination step of the propagated $O_3$ forming cycle (Figure \ref{fig:O3scheme}), and as \citep{Roberts1998} estimated it is likely to begin to occur when the $NO$ mixing ratio increases up to 100-200 ppt. The yield of alkyl nitrate formation is controlled by the branching ratio ($\alpha_3$), which is temperature and pressure dependent and increases with the number of carbons in the alkyl peroxy radical \citep{Roberts1990}.

@@

The lifetimes of alkyl nitrates depend on the seasonality of the photolysis and $OH$ sink, which in turn vary with latitude and altitude. According to \citep{Clemitshaw1997} and \citep{Turberg1990} alkyl nitrates have atmospheric lifetimes of several days to several years, which makes them temporary reservoirs for the short lived $NO_x$ compounds. Being transported from polluted areas to remote background regions and eventually destroyed alkyl nitrates can 'release' $NO_2$ back to the atmosphere, thereby contributing to tropospheric ozone formation.
\subsection{Ozone, alkyl nitrates and carbon bonds}\label{sec:intro_CB}
In the previous sections we covered the chemistry that controls the production of $O_3$ in the troposphere. A traditional way to assess the efficiency of this production is to calculate the ratio:
\begin{matriz}\label{eq:OPE}
OPE = \dfrac{k_{HO_2+NO}[HO_2][NO]}{k_{OH+NO_2}[OH][NO_2]} = \dfrac{P_{O_3}}{L_{NO_x}}
\end{matriz}
where $OPE$ is the ozone production efficiency, $k_{HO_2+NO}$ - the rate coefficient of reaction (\ref{reac:HO2+NO=NO2+OH}), $k_{OH+NO_2}$ - the rate coefficient of reaction (\ref{reac:OH+NO2=HNO3}), $[X]$ - mixing ratios of corresponding species, $P_{O_3}$ – the $O_3$ production rate, $L_{NO_x}$ – the $NO_x$ loss rate. $OPE$ is determined by the amount of $HO_2$ reacting with $NO$ to form $NO_2$ and $OH$ (i.e., $k_{HO_2+NO}[HO_2][NO]$, (\ref{reac:HO2+NO=NO2+OH})) and the amount of $NO_x$ removed from the system by reaction of $OH$ with $NO_2$ to form nitric acid ($HNO_3$, reaction (\ref{reac:OH+NO2=HNO3})). In other words, $OPE$ shows the amount of ozone created ($P_{O_3}$) per molecule of $NO_x$ removed ($L_{NO_x}$) from the system \citep{Jacob1999}. Due to the reasons described in section \ref{sec:method_chem} \footnote{No representation of the nitric acid ($HNO_3$) in the model.}, we use another expression for calculations of the $O_3$ production efficiency. This expression has been recently proposed by Professor Matthew Evans (University of York) amid the mounting concerns about the need of thorough intercomparison of chemical mechanisms of global atmospheric models (including those used in the IPCC reports), why it is more universal and applicable to any kind of model.

According to \citep{Evans2014} (personal communication), the $O_3$ production efficiency can be estimated via calculation of the number of carbon bonds emitted into the system. For example, if the system contains one molecule of methane, $CH_4$, as 'a carbon bond reservoir', it means that this system has 4 carbon bonds (4 C-H), if  instead of methane it contains one molecule of ethane, $C_2H_6$, the system has 7 carbon bonds (6 C-H bonds and 1 C-C), and so on \citep{Edwards2013}. Assuming that every carbon bond is broken chemically and every organic peroxy radical ($RO_2$) reacts with $NO$, the overall efficiency of carbon bonds conversion to $O_3$, $\epsilon$, can be calculated as: 
\begin{matriz}\label{eq:epsilon}
\epsilon = \dfrac{P_{O_3}}{E_{C_{bons}}} = \chi\cdot\upsilon
\end{matriz}
where $P_{O_3}$ – the rate of $O_3$ production (mol yr-1), $E_{C_{bons}}$  – the rate of carbon bonds emission (mol yr-1), $\chi$ – ratio between the rate of carbon bonds loss to form $RO_2$ and the loss rate of all carbon bonds, $\upsilon$ – ratio between the rate of $RO_2$ loss in reaction with $NO$ resulting in $NO$-to-$NO_2$ conversion and the loss rate of all $RO_2$. In other words, chemical loss of 1 carbon bond produces 1 peroxy radical:
\begin{itemize}
\centering
\item[] $OH + {CH_4}|_{4 bonds} \rightarrow {CH_3O_2}|_{3 bonds} + H_2O$
\item[] 4 carbon bonds = 3 carbon bonds, 1 peroxy radical
\end{itemize}
Loss of 1 carbon bond via photolysis produces 2 peroxy radicals:
\begin{itemize}
\centering
\item[] ${HCHO}|_{2bonds} + hv \rightarrow H + {HCO}|_{1bond}$
\item[] ${HCO}|_{1bond} + O_2 \rightarrow HO_2 + {CO}|_{1bond}$
\item[] $H + O_2 \xrightarrow{M} HO_2$
\item[] $horizontal line$
\item[] 2 carbon bonds = 1 carbon bond, 2 peroxy radicals
\end{itemize}
or no peroxy radicals because of dry or wet deposition:
\begin{itemize}
\centering
\item[] $HCHO \xrightarrow{deposition}$
\item[] 2 carbon bonds = 2 carbon bonds, 0 peroxy radicals
\end{itemize}

The fact that $\epsilon$ does not account for reactions where no $NO$-to-$NO_2$ conversion happens, as, for example, in case of the formation of the alkyl nitrates, $\epsilon$ provides a way to relate the rate of $O_3$ production to the alkyl nitrate chemistry. Knowing the rate of carbon bond emissions ($E_{C_{bons}}$) from emissions inventory data and knowing the rate of $O_3$ production from a model run ($P_{O_3}$), one can estimate the efficiency of $O_3$ production per carbon bond emitted in their model, and than compare it with other such values from different models potentially revealing the true cause of discrepancies between models and observations of tropospheric $O_3$ budgets.

For link at the bottom? The same kind of approach is used in the Common Representative Intermediates (CRI) mechanism for VOC degradation \citep{Jenkin2002, Jenkin2008}. It is a reduced chemical mechanism that describes $O_3$ formation from the tropospheric degradation of methane and 115 emitted non-methane hydrocarbons and oxygenated volatile organic compounds developed using the Master Chemical Mechanism version 3.1 (MCM v3.1) as a reference benchmark
\section{Overall objective and specific aims}\label{sec:objams}
As the reader might anticipateit is mentioned in section \ref{sec:lit_review}
\citep{Sommariva2008}
[Typically, chemical mechanisms do not contain
enough information about the various formation routes of secondary organic compounds, leading to the underestimation of their concentrations, with possible consequences on the estimates of ozone production in remote regions, or to ineffective ozone-control policies. In this work, a quasi-explicit chemical model based upon the Master Chemical Mechanism (MCM) has been used to study in detail the photochemistry of a number of alkyl nitrates and determine quantitatively their formation processes. This kind of study shows the potential of using mechanisms with a high level of chemical detail to study the formation of pollutants in the atmosphere (Madronich, 2006).]
\subsection{Overall objective}\label{sec:objams_obj}
The overall purpose of this research is to investigate the impact of alkyl nitrate chemistry on ozone production efficiency.
\subsubsection*{Specific aims}\label{sec:objams_ams}
Specific aims are to investigate:
\begin{itemize}
\item[(1)] the relationship between ozone production and carbon bond emissions;
\item[(2)] the impact of inclusion of alkyl nitrates chemistry on this relationship;
\item[(3)] how (1) and (2) vary in different $NO_x$ and $VOC$ regimes;
\item[(4)] build different version of the box model;
\item[(5)] to evaluate the results.
\end{itemize}

>>>>

Unlike many other air pollutants, tropospheric ozone is not directly emitted, but produced following the oxidation of carbon monoxide ($CO$), methane ($CH_4$), and nonmethane volatile compounds (nmVOCs) in the presence of nitrogen oxides ($NO_x$) \citep{Crutzen1973,Myhre2013}. The emissions of these precursors arise from both natural and anthropogenic sources which have intensified since pre-industrial times \citep{Parrish2014,Volz1988} (Figure \ref{fig:O3observations}). Presumably it has led to roughly a doubling in background tropospheric ozone concentration \citep{Guicherit2000,Hartmann2013,Tarasick2008,Vingarzan2004}, although contribution from another major source, transport from the stratosphere, may also took place \citep{Fowler2008}. Climate is also believed to be a key driver of ozone production and destruction since the chemical reactions affecting ozone concentration are strongly dependent on such climatic factors as temperature, rainfall and humidity. Therefore in a changing climate tropospheric ozone is not anymore a local air quality issue, but is a global pollution problem \citep{Fowler2008}.

@@

According to the Intergovernmental Panel on Climate Change (IPCC) Fifth Assessment report (AR5) in the present conditions (around year 2000) the global mean tropospheric ozone budget is approximately 331 Tg \citep{Myhre2013}. However, despite general agreement on how the drivers impact ozone concentrations at global and regional scales, there is considerable inconsistency in its budget between different modelling and observational studies \citep{Stevenson2006,Wild2007,Young2012}. Figure \ref{fig:O3budget} demonstrates that the range of global tropospheric ozone budget estimates is approximately 60 Tg which is 18\% of the mean. In terms of ozone production the variation is even higher and accounts for 28\% of the corresponding value. In that regard, two questions arise: why does the tropospheric ozone budget differ between models and observations? And secondly, why does it vary to this extent, especially in terms of production? The only way to answer these questions is to perform a rigorous investigation of the factors that drive tropospheric ozone and attribute those differences to respective factors \citep{Young2012}.

\begin{figure}[t] % O3 budget
\centering
\includegraphics[width=0.8\linewidth]{o3_studies.eps}
\caption{Summary of tropospheric ozone global budget model and observation estimates for present (about 2000) conditions. All uncertainties quoted as 1 standard deviation (68\% confidence interval). Adapted from \citep{Myhre2013}}\label{fig:O3budget}
\end{figure}

A significant effort has been made to tackle these problems, and broadly speaking there are two possible ways of doing it. The first one implies the direct analysis of tropospheric chemical schemes used in climate-chemistry models. For example, \cite{Emmerson2009} compared the state of the art Master Chemical Mechanism which contains approximately 5600 species and 13500 reactions \citep{Jenkin2002} with other six chemical schemes and found significant variations between them. This approach is associated with serious challenges since a modern global model is a very sophisticated representation of the climate-chemistry interactions, and therefore it is often difficult to separate influence of one reaction from another. However, even advanced climate-chemistry models do not take into account all known chemical reactions in the real troposphere. A complete representation requires many thousands of species and tens of thousands of reactions, which is beyond the numerical capabilities currently available. This significantly limits the practicable size of chemical schemes and typically involves a reduction of the number of VOCs considered and lumping the carbon from the discarded species into representative surrogates.

The second way to understanding the discrepancies between models and observations is to conduct an idealised theoretical study using a conceptual model. The advantage of this approach is that such kind of models can be quite easily constructed, and more importantly can be focused on representation of a certain chemical reactions which possibly could explain the discrepancy between models and observations.

\subsubsection*{Ozone tropospheric chemistry}\label{intro_O3chem}
Keeping in mind the role of tropospheric ozone, a necessary step is to revisit the chemistry that controls its abundance.

The key chemical processes that lead to $O_3$ formation and destruction in the troposphere are driven by reaction cycles involving free-radical intermediates, which are mainly formed from the photolysis of $O_3$ itself (\ref{reac:O3+hv=O1D+O2}) \citep{Fowler2008}. At UV wavelengths shorter than 320 nm, $O_3$ photolysis generates excited oxygen atoms ($O(^1D)$), which than can either react with water vapour to form $OH$ radicals (\ref{reac:O1D+H2O=OH+OH}), or collide with an inert molecule (denoted '$M$') to reform $O_3$ (\ref{reac:O1D=O3P} and \ref{reac:O3P+O2=O3}):
\begin{equation}\label{reac:O3+hv=O1D+O2}
O_3 + hv \rightarrow O(^1D) + O_2
\end{equation}
\begin{equation}\label{reac:O1D+H2O=OH+OH}
O(^1D) + H_2O \rightarrow OH + OH
\end{equation}
\begin{equation}\label{reac:O1D=O3P}
O(^1D) \xrightarrow{M} O(^3P)
\end{equation}
\begin{equation}\label{reac:O3P+O2=O3}
O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
Once formed the OH radicals mainly react with $CO$ and $CH_4$ and initiate the $O_3$ formation or removal cycles by producing the peroxy radicals ($HO_2$, $CH_3O_2$):
\begin{equation}\label{reac:OH+CO=H+CO2}
OH + CO \rightarrow H + CO_2
\end{equation}
\begin{equation}\label{reac:H+O2=HO2}
H + O_2 \xrightarrow{M} HO_2
\end{equation}
\begin{equation}\label{reac:OH+CH4=CH3+H2O}
OH + CH_4 \rightarrow CH_3 + H_2O
\end{equation}
\begin{equation}\label{reac:CH3O+O2=CH3O2}
CH_3 + O_2 \xrightarrow{M} CH_3O_2
\end{equation}
The fate of the peroxy radicals depends on the presence of $NO_x$ ($=NO + NO_2$). In low-$NO_x$ conditions (remote regions), the $HO_2$ radical reacts with another $HO_2$ to form hydrogen peroxide ($H_2O_2$) or reacts with the methyl peroxy radical ($CH_3O_2$) to form methyl hydrogen peroxide ($CH_3OOH$):
\begin{equation} \label{reac:HO2+HO2=H2O2+O2}
HO_2 + HO_2 \rightarrow H_2O_2 + O_2
\end{equation}
\begin{equation} \label{reac:HO2+CH3O2=CH3OOH+O2}
HO_2 + CH_3O_2 \rightarrow CH_3OOH + O_2
\end{equation}
On a whole, in low-$NO_x$ conditions, a net $O_3$ destruction takes place, because the reaction sequence is initiated by $O_3$ photolysis. Plus in the same conditions some additional $O_3$ removal can occur because of the reaction of the $HO_2$ radical with $O_3$ leading to further destruction of ozone in a chain sequence involving formation of $OH$ radicals:
\begin{equation}\label{reac:HO2+O3=OH+2O2}
HO_2 + O_3 \rightarrow OH + 2O_2
\end{equation}
\begin{equation}\label{reac:OH+O3=HO2+O2}
OH + O_3 \rightarrow HO_2 + O_2
\end{equation}

In intermediate $NO_x$ conditions (rural areas of most industrialised countries), the peroxy radicals react with nitrogen monoxide ($NO$) to form nitrogen dioxide ($NO_2$), which subsequent photolysis leads to $O_3$ generation:
\begin{equation}\label{reac:HO2+NO=NO2+OH}
HO_2 + NO \rightarrow NO_2 + OH
\end{equation}
\begin{equation}\label{reac:CH3O2+NO=NO2+CH3O}
CH_3O_2 + NO \rightarrow NO_2 + CH_3O
\end{equation}
\begin{equation}\label{reac:NO2+hv=NO+O3P}
NO_2 + hv \rightarrow NO + O(^3P)
\end{equation}
\begin{equation}
O(^3P) + O_2 \xrightarrow{M} O_3\tag{\ref{reac:O3P+O2=O3}}
\end{equation}
As shown in Figure \ref{fig:O3scheme}, the reactions (\ref{reac:HO2+NO=NO2+OH}) and (\ref{reac:CH3O2+NO=NO2+CH3O}) produce parts of the free-radical propagated $O_3$ forming cycles, which may occur a number of times before being halted by radical termination reactions (\ref{reac:HO2+HO2=H2O2+O2} and \ref{reac:HO2+CH3O2=CH3OOH+O2}), and the higher $NO_x$ level is the greater number of these cycles can occur. Therefore, in intermediate $NO_x$ conditions a net $O_3$ formation takes place and it is sensitive to $NO_x$.

@@

In high-$NO_x$ conditions (urban environment), the $OH$ radical may have another fate rather than reacting with $CO$, $CH_4$ and $O_3$. Depending on the concentration of $NO_2$ it can react with it to form nitric acid ($HNO_3$):
\begin{equation}\label{reac:OH+NO2=HNO3}
OH + NO_2 \xrightarrow{M} HNO_3
\end{equation}
The formation of $HNO_3$ terminates free-radical propagated $O_3$ forming cycles leading to a decrease in $O_3$ production rate. However, elevated emissions of $CO$ and $CH_4$ (and other nonmethane volatile organic compounds (nmVOCs)) may make reactions (\ref{reac:OH+CO=H+CO2})-(\ref{reac:OH+CH4=CH3+H2O}) more competitive with the reaction (\ref{reac:OH+NO2=HNO3}) and lead to an increase in the $O_3$ production rate. Therefore, in high-$NO_x$ conditions the $O_3$ production rate can either decrease or increase depending on $NO_x$ and VOCs levels.

@@

therefore also having an impact on the rates of $O_3$ production \citep{Fowler2008} as do some other species, which are not mentioned in this section. Their role is described with a required depth in the sections below.

\subsection{Alkyl nitrates}\label{sec:intro_AN}
Alkyl nitrates ($RONO_2$) are organic trace gases, which through one of their formation mechanisms are closely linked with the production of tropospheric ozone \citep{Reeves2007}. This mechanism is presented in the following section.
\subsubsection*{Alkyl nitrates chemistry}\label{into_ANchem}
The photochemical formation of the alkyl nitrates starts when chemistry of the troposphere reaches the regime of $O_3$ production, i.e. when reactions of the peroxy radicals with $NO$ start to compete with their self- and cross-reactions more effectively due to an increase in $NO_x$ mixing ratio. In this case the reaction between the alkyl peroxy radical ($RO_2$, where $R$ denotes methyl group, $CH_3$) and $NO$ can reveal its two-channel nature \citep{Day2003}:
\begin{subequations} \label{reac:RONO2alpha}
\begin{align}
RO_2 + NO &\xrightarrow{1-\alpha_3} RO + NO_2 \label{reac:RO2+NO=aphaRO+NO2}\\
RO_2 + NO &\xrightarrow{\alpha_3} RONO_2 \label{reac:RO2+NO=aphaRONO2}
\end{align}
\end{subequations}
The latter reaction can serve as a termination step of the propagated $O_3$ forming cycle (Figure \ref{fig:O3scheme}), and as \cite{Roberts1998} estimated it is likely to begin to occur when the $NO$ mixing ratio increases up to 100-200 ppt. The yield of alkyl nitrate formation is controlled by the branching ratio ($\alpha_3$), which is temperature and pressure dependent and increases with the number of carbons in the alkyl peroxy radical \citep{Roberts1990}.

@@

The lifetimes of alkyl nitrates depend on the seasonality of the photolysis and $OH$ sink, which in turn vary with latitude and altitude. According to \cite{Clemitshaw1997} and \cite{Turberg1990} alkyl nitrates have atmospheric lifetimes of several days to several years, which makes them temporary reservoirs for the short lived $NO_x$ compounds. Being transported from polluted areas to remote background regions and eventually destroyed alkyl nitrates can 'release' $NO_2$ back to the atmosphere, thereby contributing to tropospheric ozone formation.

\subsection{Ozone, alkyl nitrates and carbon bonds}\label{sec:intro_CB}
In the previous sections we covered the chemistry that controls the production of $O_3$ in the troposphere. A traditional way to assess the efficiency of this production is to calculate the ratio:
\begin{matriz}\label{eq:OPE}
OPE = \dfrac{k_{HO_2+NO}[HO_2][NO]}{k_{OH+NO_2}[OH][NO_2]} = \dfrac{P_{O_3}}{L_{NO_x}}
\end{matriz}
where $OPE$ is the ozone production efficiency, $k_{HO_2+NO}$ - the rate coefficient of reaction (\ref{reac:HO2+NO=NO2+OH}), $k_{OH+NO_2}$ - the rate coefficient of reaction (\ref{reac:OH+NO2=HNO3}), $[X]$ - mixing ratios of corresponding species, $P_{O_3}$ – the $O_3$ production rate, $L_{NO_x}$ – the $NO_x$ loss rate. $OPE$ is determined by the amount of $HO_2$ reacting with $NO$ to form $NO_2$ and $OH$ (i.e., $k_{HO_2+NO}[HO_2][NO]$, (\ref{reac:HO2+NO=NO2+OH})) and the amount of $NO_x$ removed from the system by reaction of $OH$ with $NO_2$ to form nitric acid ($HNO_3$, reaction (\ref{reac:OH+NO2=HNO3})). In other words, $OPE$ shows the amount of ozone created ($P_{O_3}$) per molecule of $NO_x$ removed ($L_{NO_x}$) from the system \citep{Jacob1999}. Due to the reasons mentioned in section \ref{sec:method_chem} \footnote{No representation of the nitric acid ($HNO_3$) in the model.}, we use another expression for calculations of the $O_3$ production efficiency. This expression has been recently proposed by Professor Matthew Evans (University of York) amid the mounting concerns about the need of thorough intercomparison of chemical mechanisms of global atmospheric models (including those used in the IPCC reports), why it is more universal and applicable to any kind of model.

According to \cite{Evans2014} (personal communication), the $O_3$ production efficiency can be estimated via calculation of the number of carbon bonds emitted into the system\footnote{The same kind of approach is used in the Common Representative Intermediates (CRI) mechanism for VOC degradation \citep{Jenkin2002, Jenkin2008}. It is a reduced chemical mechanism that describes $O_3$ formation from the tropospheric degradation of methane and 115 emitted non-methane hydrocarbons and oxygenated volatile organic compounds developed using the Master Chemical Mechanism version 3.1 (MCM v3.1) as a reference benchmark.}. For example, if the system contains one molecule of methane, $CH_4$, as 'a carbon bond reservoir', it means that this system has 4 carbon bonds (4 $C-H$), if  instead of methane it contains one molecule of ethane, $C_2H_6$, the system has 7 carbon bonds (6 $C-H$ bonds and 1 $C-C$), and so on \citep{Edwards2013}. Assuming that every carbon bond is broken chemically and every organic peroxy radical ($RO_2$) reacts with $NO$, the overall efficiency of carbon bonds conversion to $O_3$, $\epsilon$, can be calculated as: 
\begin{matriz}\label{eq:epsilon}
\epsilon = \dfrac{P_{O_3}}{E_{C_{bons}}} = \chi\cdot\upsilon
\end{matriz}
where $P_{O_3}$ – the rate of $O_3$ production (mol yr-1), $E_{C_{bons}}$  – the rate of carbon bonds emission (mol yr-1), $\chi$ – ratio between the rate of carbon bonds loss to form $RO_2$ and the loss rate of all carbon bonds, $\upsilon$ – ratio between the rate of $RO_2$ loss in reaction with $NO$ resulting in $NO$-to-$NO_2$ conversion and the loss rate of all $RO_2$. In other words, chemical loss of 1 carbon bond produces 1 peroxy radical:
\begin{itemize}
\centering
\item[] $OH + {CH_4}|_{4 bonds} \rightarrow {CH_3O_2}|_{3 bonds} + H_2O$
\item[] 4 carbon bonds = 3 carbon bonds, 1 peroxy radical
\end{itemize}
Loss of 1 carbon bond via photolysis produces 2 peroxy radicals:
\begin{itemize}
\centering
\item[] ${HCHO}|_{2bonds} + hv \rightarrow H + {HCO}|_{1bond}$
\item[] ${HCO}|_{1bond} + O_2 \rightarrow HO_2 + {CO}|_{1bond}$
\item[] $H + O_2 \xrightarrow{M} HO_2$
\item[] $horizontal line$
\item[] 2 carbon bonds = 1 carbon bond, 2 peroxy radicals
\end{itemize}
or no peroxy radicals because of dry or wet deposition:
\begin{itemize}
\centering
\item[] $HCHO \xrightarrow{deposition}$
\item[] 2 carbon bonds = 2 carbon bonds, 0 peroxy radicals
\end{itemize}

The fact that $\epsilon$ does not account for reactions where no $NO$-to-$NO_2$ conversion happens, as, for example, in case of the formation of the alkyl nitrates, $\epsilon$ provides a way to relate the rate of $O_3$ production to the alkyl nitrate chemistry. Knowing the rate of carbon bond emissions ($E_{C_{bons}}$) from emissions inventory data and knowing the rate of $O_3$ production from a model run ($P_{O_3}$), one can estimate the efficiency of $O_3$ production per carbon bond emitted in their model, and than compare it with other such values from different models potentially revealing the true cause of discrepancies in tropospheric $O_3$ budgets between models and observations.

\section{Overall objective and specific aims}\label{sec:objams}
Despite of all of the aforesaid, typically chemical mechanisms of global atmospheric models do not contain enough information about some secondary organic compounds \citep{Sommariva2008}, and alkyl nitrates are among them. Underestimation of their concentrations could potentially lead to errors in the estimates of ozone production in remote regions, or to ineffective ozone-control policies. Therefore, the overall objective of this research is to investigate the impact of alkyl nitrate chemistry on tropospheric ozone production by using a chemical mechanism that represents the chemistry related to the problem.

The specific aims are:
\begin{itemize}
\item[(1)] to build a conceptual box model which chemical mechanism represents the basics of tropospheric chemistry of ozone and includes reactions of formation and loss of alkyl nitrates;
\item[(2)] to estimate the sensitivity of overall tropospheric ozone production to alkyl nitrates chemistry;
\item[(3)] to explore the the relationship between ozone production and carbon bonds emissions;
\item[(3)] to estimate how (2) and (3) vary in different $NO_x$ and VOC regimes, which could be achieved by designing an appropriate set of experiments.
\end{itemize}

The literature review on the problem is presented in section \ref{sec:lit_review}. In section \ref{sec:method} the conceptual model of global troposhere is described. Section \ref{sec:res} contains the results. Section \ref{sec:discuss} discusses the results and also includes an outline of future work. Conclusions constitute section \ref{sec:conclusion}.
